Index-aware learning of circuits

### 2024-01-25 doc/article/arxiv/article.tex
graphical abstract

\usetikzlibrary{external}
\tikzexternalize

@@

\textcolor{red}{In order to obtain a version of \eqref{eq:md_mnai} that is better suited to analysis and implementation, we consider the device function Jacobians
\begin{align}
    \mb{G} \big( \AT{R} \vphi \big) \coloneqq \mb{J}_{\AT{R} \vphi}(\gR), \quad \mb{L}(\i{L}) \coloneqq \mb{J}_{\i{L}}(\phiL), \quad \mb{C} \big( \AT{C} \vphi \big) \coloneqq \mb{J}_{\AT{C} \vphi}(\qC), \label{eq:md_dfj}
\end{align}
where we use the same notation for the Jacobians as in \autoref{sec:int}.} Inserting \eqref{eq:md_dfj} into the original system, while linearizing $\gR$ and writing everything in matrix form, yields
\begin{align}
    \begin{bmatrix}
        \A{C}^{\phantom{\T}} \mb{C} \big( \AT{C} \vphi \big) \AT{C} & &\\
        & \mb{L}(\i{L}) &\\
        & & \mb{0}
    \end{bmatrix} \ddt \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{R}^{\phantom{\T}} \mb{G} \big( \AT{R} \vphi \big) \AT{R} & \A{L} & \A{V}\\
        -\AT{L} & &\\
        -\AT{V} & &
    \end{bmatrix} \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{I} \i{s}(t)\\
        \mb{0}\\
        \v{s}(t)
    \end{bmatrix} = \mb{0}, \label{eq:md_mnaii}
\end{align}
assuming that $\phiL$ and $\qC$ do not explicitly depend on time.

\textcolor{red}{We note that \eqref{eq:md_mnaii} readily extends to multiport devices\cite{tischendorf1999}. A general inductive $n$-port for example, can also be modeled by a function $\phiL(\i{L}) = \big[ \phi_1(\i{L}), \dotsc, \phi_{n-1}(\i{L}) \big]^\T$, however the Jacobian is not necessarily diagonal in this case as the component functions of $\phiL$ can each depend on all the currents $\i{L} = [i_1, \cdots, i_{n-1}]^\T$. Regarding the incidence matrices, one chooses a reference node $\varphi_0$ for the multiport device and then considers $n-1$ branches from the remaining $n-1$ nodes to the reference. Interpreting the multiport as a single node and using Kirchhoff's current law then gives $i_0 = -\sum_{k=1}^{n-1} i_k$ for the reference current, when orienting all currents to point toward the device. Following this approach, there is then no difference in treating multiports compared to one-ports\cite{tischendorf1999}.}

\subsection{Index concepts}
\label{subsec:ic}
\textcolor{red}{Before outlining the dissection index, we want to give a brief introduction to index concepts more generally. There are multiple definitions of the index of a DAE, along with related index concepts that each possess different strengths and weaknesses\cite{mehrmann2015}. One important aspect that unites these ideas is that they agree in key cases, e.g.~when looking at linear DAEs, and the same also holds true for the dissection index. To emphasize the practical importance of the notion of index, we take a closer look at the perturbation index. It is based on a perturbed version of the DAE \eqref{eq:int_dae} (we leave out the parameters $\mb{p}$ for conciseness)
\begin{align}
    \mb{F}(\mbh{x}', \mbh{x}, t) = \boldsymbol{\varepsilon}(t), \quad \mbh{x}(0) = \mbh{x}_0, \label{eq:ic_pdae}
\end{align}
where $\varepsilon$ is a sufficiently smooth perturbation, such that the required derivatives exist.
\begin{definition}
    \label{def:ic_pi}
    Let $\mb{x}$ be a solution of the unperturbed DAE, then the DAE is said to have perturbation index $\nu \in \mathbb{N}$, if $\nu$ is the smallest natural number such that, for any sufficiently smooth solution $\mbh{x}$ of \eqref{eq:ic_pdae}, there exists $c \in \mathbb{R}$ with
    \begin{align*}
        \| \mbh{x} - \mb{x} \| \leq c \big( \| \mbh{x}_0 - \mb{x}_0 \| + \| \boldsymbol{\varepsilon} \|_\infty + \| \boldsymbol{\varepsilon}' \|_\infty + \dotsb + \| \boldsymbol{\varepsilon}^{(\nu - 1)} \|_\infty \big)
    \end{align*}
    for an appropriate norm $\| \cdot \|$ and the right hand side small enough.
\end{definition}
The idea behind this definition is to capture the impact of perturbations on the solution, as the name suggests. Usually these pertubations are assumed small, in the sense that $\| \boldsymbol{\varepsilon} \|_\infty \ll 1$, but fast changing, such that $\| \boldsymbol{\varepsilon}^{(\nu)} \|_\infty$ may grow very quickly in $\nu$.}

\paragraph{Example}
\begin{figure}[b]
    \begin{center}
        \includegraphics[width=0.35\textwidth]{vcr}
    \end{center}
    \caption{Small example circuit for illustrating the perturbation index.}
    \label{fig:ic_vcr}
\end{figure}
\textcolor{red}{To illustrate the perturbation index, and also to hint at its relevance for circuit simulation, we consider the small example given in \autoref{fig:ic_vcr}. Applying MNA to the circuit and introducing a perturbation yields the following DAE
\begin{subequations}
    \label{eq:ic_vcr}
    \begin{align}
        C \ddt \hat{\varphi}_1 + \frac{1}{R} \hat{\varphi}_1 + \hat{i}_\mr{V} &= \varepsilon_1(t) \label{eq:ic_vcra}\\
        -\hat{\varphi}_1 + v_\mr{s}(t) &= \varepsilon_2(t), \label{eq:ic_vcrb}
    \end{align}
\end{subequations}
where $\mbh{x} = [\hat{\varphi}_1, \hat{i}_\mr{V}]^\T$. Using \eqref{eq:ic_vcrb} we find
\begin{align}
    \hat{\varphi}_1 = v_\mr{s}(t) - \varepsilon_2(t). \label{eq:ic_vcrphi}
\end{align}
Inserting \eqref{eq:ic_vcrphi} into \eqref{eq:ic_vcra} and rearranging then gives
\begin{align*}
    \hat{i}_\mr{V} &= \varepsilon_1(t) - C \ddt \big( v_\mr{s}(t) - \varepsilon_2(t) \big) - \frac{1}{R} \big( v_\mr{s}(t) - \varepsilon_2(t) \big).
\end{align*}
Noting that the solution $\mb{x}$ to the unperturbed problem follows directly from the perturbed solution by setting $\boldsymbol{\varepsilon} = \mb{0}$, we obtain
\begin{align}
    \| \mbh{x} - \mb{x} \| = \begin{Vmatrix}
        -\varepsilon_2(t)\\
        \varepsilon_1(t) + C \ddt \varepsilon_2(t) + \frac{1}{R} \varepsilon_2(t)
    \end{Vmatrix}, \label{eq:ic_vcrpi}
\end{align}
so the unperturbed system corresponding to \eqref{eq:ic_vcr} has perturbation index $\nu = 2$ as \eqref{eq:ic_vcrpi} depends on the first derivative of $\varepsilon_2$.}

@@

\section{Dissection index}
\label{sec:di}
\textcolor{red}{We focus on the dissection index, as it enables the decoupling of a DAE into an ODE and a set of purely algebraic equations. This is conceptually different from the perturbation index, however other index concepts such as the tractability and differentiation indices also use decoupling strategies. Still, the dissection index maintains some advantages over these concepts, as it provides a simple algorithmic procedure for the decoupling that is similar to the tractability index, but poses less strict smoothness assumptions. In the case of MNA without controlled sources \eqref{eq:md_mnaii}, it is even possible to find a purely topological decoupling based on the dissection index\cite{jansen2014}. We will not make use of this topological decoupling in the derivation however, but rather consider the dissection index for a more general class of DAEs, to formulate the assumptions that are necessary for our method to work also for DAEs other than \eqref{eq:md_mnaii}.}

\textcolor{red}{We consider a DAE in standard form\cite{jansen2014},
\begin{align}
    \mb{M}(\mb{x}) \ddt \mb{x} + \mb{K}(\mb{x}) \mb{x} + \mb{f}(t) = \mb{0}, \quad \ker \mb{M} \supsetneq \{ \mb{0} \}, \label{eq:di_dae}
\end{align}
where the matrix-valued functions $\mb{M}$ and $\mb{K}$ are derived from \eqref{eq:int_dae} by defining
\begin{align*}
    \mb{M}(\mb{x}) = \mb{J}_{\mb{x}'} \big( \mb{F}(\mb{x}', \mb{x}, t) \big), \quad \mb{K}(\mb{x}) = \mb{J}_{\mb{x}} \big( \mb{F}(\mb{x}', \mb{x}, t) \big).
\end{align*}
Note that the description of MNA in \eqref{eq:md_mnaii} is precisely of this form. In the following, we will demonstrate the first two steps of the dissection index when applied to systems of the form of \eqref{eq:di_dae}, while stating the assumptions of our approach. Appendix \autoref{sec:dim} contains additional remarks showing that these assumptions are fulfilled by \eqref{eq:md_mnaii}.}

\subsection{Index one case}
\label{subsec:ioc}
Assuming $\mb{M}(\mb{x})$ to be sufficiently smooth with values in $\mathbb{R}^{n \times m}$, we define four basis functions $\mb{P}(\mb{x})$, $\mb{Q}(\mb{x})$, $\mb{V}(\mb{x})$, $\mb{W}(\mb{x})$ such that
\begin{align*}
    \mr{im\, } \mb{Q}(\mb{x}) = \ker \mb{M}(\mb{x}), \quad \mr{im\, } \mb{W}(\mb{x}) = \ker \mb{M}^\T (\mb{x})
\end{align*}
and the columns of $\mb{P}(\mb{x})$ and $\mb{Q}(\mb{x})$ together form a basis of $\mathbb{R}^n$, while the columns of $\mb{V}(\mb{x})$ and $\mb{W}(\mb{x})$ together form a basis of $\mathbb{R}^m$. We now make the following assumption.
\textcolor{red}{\begin{assumption}
    \label{as:ioc_bf}
    The basis functions $\mb{P}$ and $\mb{Q}$ of $\mb{M}(\mb{x})$ are constant.
\end{assumption}
\autoref{as:ioc_bf} may seem restrictive, but it is fulfilled by many systems occurring in practical applications\cite{jansen2014}. This in particular includes MNA, as \autoref{re:dim_bf} shows.} The key idea of the dissection index is to use the basis functions to split the solution variables $\mb{x}$ into two parts
\begin{align}
    \mb{x} = \mb{P} \mbt{x} + \mb{Q} \mbb{x}, \label{eq:ioc_x}
\end{align}
where $\tilde{\cdot}$ is used to indicate differential (dynamic) variables and $\bar{\cdot}$ signifies algebraic (fixed) variables. When inserting the splitting \eqref{eq:ioc_x} into \eqref{eq:di_dae} we obtain
\begin{align}
    \mb{M}(\mb{x}) \mb{P} \ddt \mbt{x} + \mb{K}(\mb{x}) \mb{P} \mbt{x} + \mb{K}(\mb{x}) \mb{Q} \mbb{x} + \mb{f}(t) = \mb{0} \label{eq:ioc_dae}
\end{align}
and this motivates the notation, as only $\mbt{x}$ appears differentiated in time. The procedure then continues by multiplying \eqref{eq:ioc_dae} once with $\mb{V}^\T (\mb{x})$ and $\mb{W}^\T (\mb{x})$ each from the left, to also split the system. This yields
\begin{subequations}
    \label{eq:ioc_io}
    \begin{align}
            \underbrace{\mb{V}^\T (\mb{x}) \mb{M}(\mb{x}) \mb{P}}_{\eqqcolon \mbt{M}(\mb{x})} \ddt \mbt{x} + \underbrace{\mb{V}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{P}}_{\eqqcolon \mbt{K}_\mb{P}(\mb{x})} \mbt{x} + \underbrace{\mb{V}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{Q}}_{\eqqcolon \mbt{K}_\mb{Q}(\mb{x})} \mbb{x} + \underbrace{\mb{V}^\T (\mb{x}) \mb{f}(t)}_{\eqqcolon \mbt{f}(t)} &= \mb{0} \label{eq:ioc_ioa}\\
            \underbrace{\mb{W}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{P}}_{\eqqcolon \mbb{K}_\mb{P}(\mb{x})} \mbt{x} + \underbrace{\mb{W}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{Q}}_{\eqqcolon \mbb{K}_\mb{Q}(\mb{x})} \mbb{x} + \underbrace{\mb{W}^\T (\mb{x}) \mb{f}(t)}_{\eqqcolon \mbb{f}(t)} &= \mb{0}, \label{eq:ioc_iob}
    \end{align}
\end{subequations}
where we also introduce shorthands for the arising products. Using the fact that $\mbt{M}(\mb{x})$ is regular by construction\cite{jansen2014}, we can now define the index one case of the dissection index.
\textcolor{red}{\begin{definition}
    \label{def:ioc_io}
    The DAE \eqref{eq:di_dae} has dissection index $\nu = 1$ if $\mbb{K}_\mb{Q}(\mb{x})$ is regular.
\end{definition}}
This is motivated by the observation that \eqref{eq:ioc_iob} is a purely algebraic equation with a locally unique solution for $\mbb{x}$ in terms of $\mbt{x}$, given that $\mbb{K}_\mb{Q}(\mb{x})$ is regular. In this case \eqref{eq:ioc_ioa} then describes an ODE in the differential variables $\mbt{x}$.

\subsection{Index two case}
\label{subsec:itc}
As the example from \autoref{fig:ic_vcr} illustrates, there are many DAEs, including those described by MNA, which can have an index higher than one. In these cases the dissection index proceeds by introducing additional basis functions and continuing the splitting process in a similar fashion. We begin by focusing on the algebraic equation \eqref{eq:ioc_iob}, and consider basis functions $\mbb{P}(\mb{x})$, $\mbb{Q}(\mb{x})$, $\mbb{V}(\mb{x})$, $\mbb{W}(\mb{x})$ of $\mbb{K}_\mb{Q}(\mb{x})$, defined analogous to the ones for $\mb{M}(\mb{x})$. This allows us to further split the algebraic variables $\mbb{x}$ as follows
\begin{align}
    \mbb{x} = \mbb{P}(\mb{x}) \mbb{x}_\mb{P} + \mbb{Q}(\mb{x}) \mbb{x}_\mb{Q}. \label{eq:itc_xb}
\end{align}
Inserting this splitting into \eqref{eq:ioc_iob}, and multiplying once by $\mbb{V}^\T (\mb{x})$ and $\mbb{W}^\T (\mb{x})$ each from the left, splits the algebraic equation into two parts
\begin{subequations}
    \label{eq:itc_iti}
    \begin{align}
        \mbb{V}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{x} + \mbb{V}^\T (\mb{x}) \mbb{K}_\mb{Q}(\mb{x}) \mbb{P}(\mb{x}) \mbb{x}_\mb{P} + \mbb{V}^\T (\mb{x}) \mbb{f}(t) &= \mb{0} \label{eq:itc_itia}\\
        \mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{x} + \mbb{W}^\T (\mb{x}) \mbb{f}(t) &= \mb{0}. \label{eq:itc_itib}
    \end{align}
\end{subequations}
We can now also split the differential variables $\mbt{x}$ further, by using basis functions $\mbt{P}(\mb{x})$ and $\mbt{Q}(\mb{x})$ of $\mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x})$
\begin{align}
    \mbt{x} = \mbt{P}(\mb{x}) \mbt{x}_\mb{P} +
    \mbt{Q}(\mb{x}) \mbt{x}_\mb{Q}, \label{eq:itc_xt}
\end{align}
and inserting this splitting into the second algebraic equation \eqref{eq:itc_itib} yields
\begin{align}
    \mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{P}(\mb{x}) \mbt{x}_\mb{P} + \mbb{W}^\T (\mb{x}) \mbb{f}(t) &= \mb{0}. \label{eq:itc_xtp}
\end{align}
We now make the following two assumptions.
\textcolor{red}{\begin{assumption}
    \label{as:itc_xtpdi}
    The matrix product $\mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{P}(\mb{x}) \mbt{x}_\mb{P}$ is regular.
\end{assumption}
\begin{assumption}
    \label{as:itc_xtp}
    The basis functions $\mbt{P}$ and $\mbt{Q}$ of $\mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x})$ are constant and \eqref{eq:itc_xtp} possesses a locally unique solution for $\mbt{x}_\mb{P}$ in terms of $\mbt{x}_\mb{Q}$ and $t$.
\end{assumption}
We note that while \autoref{as:itc_xtpdi} is equivalent to the DAE not being underdetermined\cite{jansen2014}, \autoref{as:itc_xtp} is rather important for the implementation, but not for the dissection index itself. In fact our approach still works if the basis functions $\mbt{P}(\mbt{x}_\mb{Q})$ and $\mbt{Q}(\mbt{x}_\mb{Q})$ depend on $\mbt{x}_\mb{Q}$, however we focus on the stronger assumption here, since it shortens the expressions in the following without impacting the general idea and \autoref{re:dim_xtp} shows that MNA fulfills an even stronger condition than \autoref{as:itc_xtp}.}

>>>>

\usetikzlibrary{external}
% \tikzexternalize

@@

In order to obtain a version of \eqref{eq:md_mnai} that is better suited to analysis and implementation, we consider the device function Jacobians
\begin{align}
    \mb{G} \big( \AT{R} \vphi \big) \coloneqq \mb{J}_{\AT{R} \vphi}(\gR), \quad \mb{L}(\i{L}) \coloneqq \mb{J}_{\i{L}}(\phiL), \quad \mb{C} \big( \AT{C} \vphi \big) \coloneqq \mb{J}_{\AT{C} \vphi}(\qC), \label{eq:md_dfj}
\end{align}
where we use the same notation for the Jacobians as in \autoref{sec:int}. Inserting \eqref{eq:md_dfj} into the original system, while linearizing $\gR$ and writing everything in matrix form, yields
\begin{align}
    \begin{bmatrix}
        \A{C}^{\phantom{\T}} \mb{C} \big( \AT{C} \vphi \big) \AT{C} & &\\
        & \mb{L}(\i{L}) &\\
        & & \mb{0}
    \end{bmatrix} \ddt \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{R}^{\phantom{\T}} \mb{G} \big( \AT{R} \vphi \big) \AT{R} & \A{L} & \A{V}\\
        -\AT{L} & &\\
        -\AT{V} & &
    \end{bmatrix} \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{I} \i{s}(t)\\
        \mb{0}\\
        \v{s}(t)
    \end{bmatrix} = \mb{0}, \label{eq:md_mnaii}
\end{align}
assuming that $\phiL$ and $\qC$ do not explicitly depend on time.

We note that \eqref{eq:md_mnaii} readily extends to multiport devices\cite{tischendorf1999}. A general inductive $n$-port for example, can also be modeled by a function $\phiL(\i{L}) = \big[ \phi_1(\i{L}), \dotsc, \phi_{n-1}(\i{L}) \big]^\T$, however the Jacobian is not necessarily diagonal in this case as the component functions of $\phiL$ can each depend on all the currents $\i{L} = [i_1, \cdots, i_{n-1}]^\T$. Regarding the incidence matrices, one chooses a reference node $\varphi_0$ for the multiport device and then considers $n-1$ branches from the remaining $n-1$ nodes to the reference. Interpreting the multiport as a single node and using Kirchhoff's current law then gives $i_0 = -\sum_{k=1}^{n-1} i_k$ for the reference current, when orienting all currents to point toward the device. Following this approach, there is then no difference in treating multiports compared to one-ports\cite{tischendorf1999}.

\subsection{Index concepts}
\label{subsec:ic}
Before outlining the dissection index, we want to give a brief introduction to index concepts more generally. There are multiple definitions of the index of a DAE, along with related index concepts that each possess different strengths and weaknesses\cite{mehrmann2015}. One important aspect that unites these ideas is that they agree in key cases, e.g.~when looking at linear DAEs, and the same also holds true for the dissection index. To emphasize the practical importance of the notion of index, we take a closer look at the perturbation index. It is based on a perturbed version of the DAE \eqref{eq:int_dae} (we leave out the parameters $\mb{p}$ for conciseness)
\begin{align}
    \mb{F}(\mbh{x}', \mbh{x}, t) = \boldsymbol{\varepsilon}(t), \quad \mbh{x}(0) = \mbh{x}_0, \label{eq:ic_pdae}
\end{align}
where $\varepsilon$ is a sufficiently smooth perturbation, such that the required derivatives exist.
\begin{definition}
    \label{def:ic_pi}
    Let $\mb{x}$ be a solution of the unperturbed DAE, then the DAE is said to have perturbation index $\nu \in \mathbb{N}$, if $\nu$ is the smallest natural number such that, for any sufficiently smooth solution $\mbh{x}$ of \eqref{eq:ic_pdae}, there exists $c \in \mathbb{R}$ with
    \begin{align*}
        \| \mbh{x} - \mb{x} \| \leq c \big( \| \mbh{x}_0 - \mb{x}_0 \| + \| \boldsymbol{\varepsilon} \|_\infty + \| \boldsymbol{\varepsilon}' \|_\infty + \dotsb + \| \boldsymbol{\varepsilon}^{(\nu - 1)} \|_\infty \big)
    \end{align*}
    for an appropriate norm $\| \cdot \|$ and the right hand side small enough.
\end{definition}
The idea behind this definition is to capture the impact of perturbations on the solution, as the name suggests. Usually these pertubations are assumed small, in the sense that $\| \boldsymbol{\varepsilon} \|_\infty \ll 1$, but fast changing, such that $\| \boldsymbol{\varepsilon}^{(\nu)} \|_\infty$ may grow very quickly in $\nu$.

\paragraph{Example}
\begin{figure}[b]
    \begin{center}
        \includegraphics[width=0.35\textwidth]{vcr}
    \end{center}
    \caption{Small example circuit for illustrating the perturbation index.}
    \label{fig:ic_vcr}
\end{figure}
To illustrate the perturbation index, and also to hint at its relevance for circuit simulation, we consider the small example given in \autoref{fig:ic_vcr}. Applying MNA to the circuit and introducing a perturbation yields the following DAE
\begin{subequations}
    \label{eq:ic_vcr}
    \begin{align}
        C \ddt \hat{\varphi}_1 + \frac{1}{R} \hat{\varphi}_1 + \hat{i}_\mr{V} &= \varepsilon_1(t) \label{eq:ic_vcra}\\
        -\hat{\varphi}_1 + v_\mr{s}(t) &= \varepsilon_2(t), \label{eq:ic_vcrb}
    \end{align}
\end{subequations}
where $\mbh{x} = [\hat{\varphi}_1, \hat{i}_\mr{V}]^\T$. Using \eqref{eq:ic_vcrb} we find
\begin{align}
    \hat{\varphi}_1 = v_\mr{s}(t) - \varepsilon_2(t). \label{eq:ic_vcrphi}
\end{align}
Inserting \eqref{eq:ic_vcrphi} into \eqref{eq:ic_vcra} and rearranging then gives
\begin{align*}
    \hat{i}_\mr{V} &= \varepsilon_1(t) - C \ddt \big( v_\mr{s}(t) - \varepsilon_2(t) \big) - \frac{1}{R} \big( v_\mr{s}(t) - \varepsilon_2(t) \big).
\end{align*}
Noting that the solution $\mb{x}$ to the unperturbed problem follows directly from the perturbed solution by setting $\boldsymbol{\varepsilon} = \mb{0}$, we obtain
\begin{align}
    \| \mbh{x} - \mb{x} \| = \begin{Vmatrix}
        -\varepsilon_2(t)\\
        \varepsilon_1(t) + C \ddt \varepsilon_2(t) + \frac{1}{R} \varepsilon_2(t)
    \end{Vmatrix}, \label{eq:ic_vcrpi}
\end{align}
so the unperturbed system corresponding to \eqref{eq:ic_vcr} has perturbation index $\nu = 2$ as \eqref{eq:ic_vcrpi} depends on the first derivative of $\varepsilon_2$.

@@

\section{Dissection index}
\label{sec:di}
We focus on the dissection index, as it enables the decoupling of a DAE into an ODE and a set of purely algebraic equations. This is conceptually different from the perturbation index, however other index concepts such as the tractability and differentiation indices also use decoupling strategies. Still, the dissection index maintains some advantages over these concepts, as it provides a simple algorithmic procedure for the decoupling that is similar to the tractability index, but poses less strict smoothness assumptions. In the case of MNA without controlled sources \eqref{eq:md_mnaii}, it is even possible to find a purely topological decoupling based on the dissection index\cite{jansen2014}. We will not make use of this topological decoupling in the derivation however, but rather consider the dissection index for a more general class of DAEs, to formulate the assumptions that are necessary for our method to work also for DAEs other than \eqref{eq:md_mnaii}.

We consider a DAE in standard form\cite{jansen2014},
\begin{align}
    \mb{M}(\mb{x}) \ddt \mb{x} + \mb{K}(\mb{x}) \mb{x} + \mb{f}(t) = \mb{0}, \quad \ker \mb{M} \supsetneq \{ \mb{0} \}, \label{eq:di_dae}
\end{align}
where the matrix-valued functions $\mb{M}$ and $\mb{K}$ are derived from \eqref{eq:int_dae} by defining
\begin{align*}
    \mb{M}(\mb{x}) = \mb{J}_{\mb{x}'} \big( \mb{F}(\mb{x}', \mb{x}, t) \big), \quad \mb{K}(\mb{x}) = \mb{J}_{\mb{x}} \big( \mb{F}(\mb{x}', \mb{x}, t) \big).
\end{align*}
Note that the description of MNA in \eqref{eq:md_mnaii} is precisely of this form. In the following, we will demonstrate the first two steps of the dissection index when applied to systems of the form of \eqref{eq:di_dae}, while stating the assumptions of our approach. Appendix \autoref{sec:dim} contains additional remarks showing that these assumptions are fulfilled by \eqref{eq:md_mnaii}.

\subsection{Index one case}
\label{subsec:ioc}
Assuming $\mb{M}(\mb{x})$ to be sufficiently smooth with values in $\mathbb{R}^{n \times m}$, we define four basis functions $\mb{P}(\mb{x})$, $\mb{Q}(\mb{x})$, $\mb{V}(\mb{x})$, $\mb{W}(\mb{x})$ such that
\begin{align*}
    \mr{im\, } \mb{Q}(\mb{x}) = \ker \mb{M}(\mb{x}), \quad \mr{im\, } \mb{W}(\mb{x}) = \ker \mb{M}^\T (\mb{x})
\end{align*}
and the columns of $\mb{P}(\mb{x})$ and $\mb{Q}(\mb{x})$ together form a basis of $\mathbb{R}^n$, while the columns of $\mb{V}(\mb{x})$ and $\mb{W}(\mb{x})$ together form a basis of $\mathbb{R}^m$. We now make the following assumption.
\begin{assumption}
    \label{as:ioc_bf}
    The basis functions $\mb{P}$ and $\mb{Q}$ of $\mb{M}(\mb{x})$ are constant.
\end{assumption}
\autoref{as:ioc_bf} may seem restrictive, but it is fulfilled by many systems occurring in practical applications\cite{jansen2014}. This in particular includes MNA, as \autoref{re:dim_bf} shows. The key idea of the dissection index is to use the basis functions to split the solution variables $\mb{x}$ into two parts
\begin{align}
    \mb{x} = \mb{P} \mbt{x} + \mb{Q} \mbb{x}, \label{eq:ioc_x}
\end{align}
where $\tilde{\cdot}$ is used to indicate differential (dynamic) variables and $\bar{\cdot}$ signifies algebraic (fixed) variables. When inserting the splitting \eqref{eq:ioc_x} into \eqref{eq:di_dae} we obtain
\begin{align}
    \mb{M}(\mb{x}) \mb{P} \ddt \mbt{x} + \mb{K}(\mb{x}) \mb{P} \mbt{x} + \mb{K}(\mb{x}) \mb{Q} \mbb{x} + \mb{f}(t) = \mb{0} \label{eq:ioc_dae}
\end{align}
and this motivates the notation, as only $\mbt{x}$ appears differentiated in time. The procedure then continues by multiplying \eqref{eq:ioc_dae} once with $\mb{V}^\T (\mb{x})$ and $\mb{W}^\T (\mb{x})$ each from the left, to also split the system. This yields
\begin{subequations}
    \label{eq:ioc_io}
    \begin{align}
            \underbrace{\mb{V}^\T (\mb{x}) \mb{M}(\mb{x}) \mb{P}}_{\eqqcolon \mbt{M}(\mb{x})} \ddt \mbt{x} + \underbrace{\mb{V}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{P}}_{\eqqcolon \mbt{K}_\mb{P}(\mb{x})} \mbt{x} + \underbrace{\mb{V}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{Q}}_{\eqqcolon \mbt{K}_\mb{Q}(\mb{x})} \mbb{x} + \underbrace{\mb{V}^\T (\mb{x}) \mb{f}(t)}_{\eqqcolon \mbt{f}(t)} &= \mb{0} \label{eq:ioc_ioa}\\
            \underbrace{\mb{W}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{P}}_{\eqqcolon \mbb{K}_\mb{P}(\mb{x})} \mbt{x} + \underbrace{\mb{W}^\T (\mb{x}) \mb{K}(\mb{x}) \mb{Q}}_{\eqqcolon \mbb{K}_\mb{Q}(\mb{x})} \mbb{x} + \underbrace{\mb{W}^\T (\mb{x}) \mb{f}(t)}_{\eqqcolon \mbb{f}(t)} &= \mb{0}, \label{eq:ioc_iob}
    \end{align}
\end{subequations}
where we also introduce shorthands for the arising products. Using the fact that $\mbt{M}(\mb{x})$ is regular by construction\cite{jansen2014}, we can now define the index one case of the dissection index.
\begin{definition}
    \label{def:ioc_io}
    The DAE \eqref{eq:di_dae} has dissection index $\nu = 1$ if $\mbb{K}_\mb{Q}(\mb{x})$ is regular.
\end{definition}
This is motivated by the observation that \eqref{eq:ioc_iob} is a purely algebraic equation with a locally unique solution for $\mbb{x}$ in terms of $\mbt{x}$, given that $\mbb{K}_\mb{Q}(\mb{x})$ is regular. In this case \eqref{eq:ioc_ioa} then describes an ODE in the differential variables $\mbt{x}$.

\subsection{Index two case}
\label{subsec:itc}
As the example from \autoref{fig:ic_vcr} illustrates, there are many DAEs, including those described by MNA, which can have an index higher than one. In these cases the dissection index proceeds by introducing additional basis functions and continuing the splitting process in a similar fashion. We begin by focusing on the algebraic equation \eqref{eq:ioc_iob}, and consider basis functions $\mbb{P}(\mb{x})$, $\mbb{Q}(\mb{x})$, $\mbb{V}(\mb{x})$, $\mbb{W}(\mb{x})$ of $\mbb{K}_\mb{Q}(\mb{x})$, defined analogous to the ones for $\mb{M}(\mb{x})$. This allows us to further split the algebraic variables $\mbb{x}$ as follows
\begin{align}
    \mbb{x} = \mbb{P}(\mb{x}) \mbb{x}_\mb{P} + \mbb{Q}(\mb{x}) \mbb{x}_\mb{Q}. \label{eq:itc_xb}
\end{align}
Inserting this splitting into \eqref{eq:ioc_iob}, and multiplying once by $\mbb{V}^\T (\mb{x})$ and $\mbb{W}^\T (\mb{x})$ each from the left, splits the algebraic equation into two parts
\begin{subequations}
    \label{eq:itc_iti}
    \begin{align}
        \mbb{V}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{x} + \mbb{V}^\T (\mb{x}) \mbb{K}_\mb{Q}(\mb{x}) \mbb{P}(\mb{x}) \mbb{x}_\mb{P} + \mbb{V}^\T (\mb{x}) \mbb{f}(t) &= \mb{0} \label{eq:itc_itia}\\
        \mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{x} + \mbb{W}^\T (\mb{x}) \mbb{f}(t) &= \mb{0}. \label{eq:itc_itib}
    \end{align}
\end{subequations}
We can now also split the differential variables $\mbt{x}$ further, by using basis functions $\mbt{P}(\mb{x})$ and $\mbt{Q}(\mb{x})$ of $\mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x})$
\begin{align}
    \mbt{x} = \mbt{P}(\mb{x}) \mbt{x}_\mb{P} +
    \mbt{Q}(\mb{x}) \mbt{x}_\mb{Q}, \label{eq:itc_xt}
\end{align}
and inserting this splitting into the second algebraic equation \eqref{eq:itc_itib} yields
\begin{align}
    \mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{P}(\mb{x}) \mbt{x}_\mb{P} + \mbb{W}^\T (\mb{x}) \mbb{f}(t) &= \mb{0}. \label{eq:itc_xtp}
\end{align}
We now make the following two assumptions.
\begin{assumption}
    \label{as:itc_xtpdi}
    The matrix product $\mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x}) \mbt{P}(\mb{x}) \mbt{x}_\mb{P}$ is regular.
\end{assumption}
\begin{assumption}
    \label{as:itc_xtp}
    The basis functions $\mbt{P}$ and $\mbt{Q}$ of $\mbb{W}^\T (\mb{x}) \mbb{K}_\mb{P}(\mb{x})$ are constant and \eqref{eq:itc_xtp} possesses a locally unique solution for $\mbt{x}_\mb{P}$ in terms of $\mbt{x}_\mb{Q}$ and $t$.
\end{assumption}
We note that while \autoref{as:itc_xtpdi} is equivalent to the DAE not being underdetermined\cite{jansen2014}, \autoref{as:itc_xtp} is rather important for the implementation, but not for the dissection index itself. In fact our approach still works if the basis functions $\mbt{P}(\mbt{x}_\mb{Q})$ and $\mbt{Q}(\mbt{x}_\mb{Q})$ depend on $\mbt{x}_\mb{Q}$, however we focus on the stronger assumption here, since it shortens the expressions in the following without impacting the general idea and \autoref{re:dim_xtp} shows that MNA fulfills an even stronger condition than \autoref{as:itc_xtp}.
